\section*{Foreword}\label{sec:foreword}
\addcontentsline{toc}{chapter}{\nameref{sec:foreword}}

Because this book is long and has never been professionally edited, it suffers from the kinds of typos and errata that plague any work of its size.  That said, I have taken some pains to be certain that the information is accurate.  Still, errors do creep in.

Corrections in the Second Edition:
\begin{enumerate}
\item Equation 6.9 originally read $p=3mn\left<u^2\right>$, which is an error. It now reads $p=\frac{1}{3}mn\left<u^2\right>$.
\end{enumerate}

Changes in the Third Edition:
\begin{enumerate}
\item Originally, the entropy of mixing was not calculated in the \texttt{igmix} class, and it was given no discussion in this handbook. The treatment in Chapter \ref{ch:ig} (specifically Sections \ref{sec:ig:mix} and \ref{sec:ig:igmix} now correctly treats the entropy of mixing.
\item $f$ was adopted for Helmholtz free energy.  In Chapter \ref{ch:mp}, $a$ was replaced with $f$. Added Section \ref{sec:intro:f}.
\item $g$ was adopted for Gibbs energy.  Added Section \ref{sec:intro:g}.
\item $a$ was adopted for speed of sound.  Added Section \ref{sec:intro:a}.
\item A number of minor corrections (e.g. typos, grammar, style) were made throughout.
\end{enumerate}

\section*{Nomenclature}\label{sec:nom}
\addcontentsline{toc}{chapter}{\nameref{sec:nom}}

These variables and units are restricted to derivations in this document.  Because \PM\ has a user-configurable unit system, any of these may be expressed in alternate units.  

The following are parameters with units.

\vspace{1em}
\begin{tabular}{|ccl|}
\hline
Symb. & Units & Description\\
\hline
$a$ & m/s & Speed of sound\\
$c$ & J/kg/K & Specific heat\\
$e$ & J/kg & Internal energy\\
$f$ & J/kg & Free energy\\
$g$ & J/kg & Gibbs energy\\
$h$ & J/kg & Enthalpy\\
$k$ & J/K & Boltzmann constant\\
$M$ & kg & Mass\\
$N$ & count & Number of moles\\
$n$ & m$^{-3}$ & Number density\\
$p$ & Pa & Pressure\\
$q$ & J kg$^{-1}$ & Normalized heat addition\\
$R$ & J kg$^{-1}$ K$^{-1}$ & Gas constant\\
$R_u$ & J kmol$^{-1}$ K$^{-1}$ & Universal gas const.\\
$s$ & J/kg/K & Entropy\\
$T$ & K & Temperature\\
$v$ & m$^3$/kg & Specific volume\\
$W$ & kg/kmol & Molar mass\\
$\rho$ & kg/m$^3$ & Density\\
\hline
\end{tabular}
\vspace{1em}

The following are dimensionless parameters.

\vspace{1em}
\begin{tabular}{|ccl|}
\hline
Symb. & Def. & Description\\
\hline
$\alpha$ & $f / RT$ & ND Free energy\\
$\delta$ & $\rho/\rho_c$ & ND Density\\
$\tau$ & $T_c / T$ & ND Temperature\\
$\gamma$ & $c_p / c_v$ & Sp. heat ratio\\
$y_i$ & $m_i / m$ & Mass fraction\\
$\chi_i$ & $N_i / N$ & Mole fraction\\
$\pi$ & $p / p_c$ & ND Pressure\\
\hline
\end{tabular}
\vspace{1em}

The following are subscripts used throughout the document.

\vspace{1em}
\begin{tabular}{|cl|}
\hline
Symb. & Description\\
\hline
$x_c$ & Critical point\\
$x_f$ & Formation property (e.g. enthalpy of formation)\\
$x_p$ & Constant-pressure (sp.heat)\\
$x_t$ & Triple point\\
$x_v$ & Constant-volume (sp.heat)\\
$x^\circ$ & Property at reference pressure\\
$\overline{x}$ & Property in molar or number units\\
\hline
\end{tabular}

Special constants used in more fundamental calcuations are defined throughout this document.  Many are listed explicitly in Table \ref{tab:constants} in Chapter \ref{ch:units}.
